---
title: "p8105_hw2_jm5509"
author: "Echo"
date: "2022-09-28"
output: github_document
---
## Problem 1
```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

```{r read and clean data}
nyc_df <- read_csv('NYC_Transit_Subway_Entrance_And_Exit_Data.csv',
                   col_types=cols(
                     `Route8`=col_character(),
                     `Route9`=col_character(),
                     `Route10`=col_character(),
                     `Route11`=col_character(),
                   )) %>% 
  janitor::clean_names() %>% 
  select(line,station_name,station_latitude,station_longitude,
         starts_with('route'),entry, vending, entrance_type, ada) %>% 
  mutate(entry=recode(entry, 'YES'=TRUE,'NO'=FALSE))

nyc_df
```
This is the first step of data management——reading and cleaning the data. This 
nyc transit data include the information of the nyc subway station names, the 
entries and exits, the location of the stations, the route that go pass the 
station, and the vending facilities. I clean the data by setting some of the 
numeric variables(route8:11) as character variables, and clean all of the 
variables names as the combination of lower-case letters and underscores. Then I
select some of the variables I would probably use later, and change the value of
"entry" variable from "YES"/"NO" into the logical value of TRUE/FALSE.

**However, these data are still not tidy now.** These route1 to route11 
variables spread across columns, which correspond to four observation times. We
can do the data-tidy job steo by step while answering the following questions.

**How many distinct stations are there?**
To get the information of the stations, we need to get the distinct station data.
```{r distinct station}
(nyc_df_distinct_station <- nyc_df%>% 
  distinct(line,station_name,.keep_all = TRUE))
nrow(nyc_df_distinct_station)
```
Identified by line and station name, we know that **465** stations are distinct.

**How many stations are ADA compliant?**
```{r ADA compliant}
(nyc_df_distinct_station %>% filter(ada==TRUE) %>% nrow())
```
Of all the distinct stations, **84** stations are ADA-compliant.

**What proportion of station entrances / exits without vending allow entrance?**
```{r allow entrance proportion}
(entry_count <- nyc_df %>% 
  select(line,station_name,entry,vending) %>% 
  filter(vending=='NO',entry==TRUE) %>% 
  nrow())

(novending_count <- nyc_df %>% 
 select(line,station_name,entry,vending) %>% 
  filter(vending=='NO') %>% 
  nrow())

(proportion <- entry_count/novending_count)
```
**37.7%** of station entrances / exits without vending allow entrance. There are 
183 entrances / exits without vending, and 69 of these allow entry.

Here, before answering the next question, we can fix the route variable problem
talked about before by using the pivot_longer().
```{r reformat data}
(nyc_df_reformat <-  nyc_df %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_name", 
    values_to = "route_number"))
```
In this case, the route number and route name are distinct variables. We can 
continue working on the following questions.

**How many distinct stations serve the A train?**
```{r A train}
(A_count <- nyc_df_reformat %>% 
  filter(route_number=='A') %>% 
  distinct(line,station_name,.keep_all = TRUE) %>% 
  nrow())
```
**60** is the number of the stations that serve the A train.

**Of the stations that serve the A train, how many are ADA compliant?**
```{r A and ADA compliant}
A_ADAcompliant_count <-nyc_df_reformat %>% 
  filter(route_number=='A',ada==TRUE) %>% 
  distinct(line,station_name,.keep_all = TRUE) %>% 
  nrow() 
```
**17** is the number of the ADA-compliant stations that serve the A train.

## Problem 2
This problem is about reading and tidying trash data. First I will read the 
corresponding sheet of the excel file, and then omit the rows and columns that 
do not contain the trash data. Furthermore, I will rename the first column and
filter the rows that contain NA and the total number. In the end, I will round 
the number of sports balls to the nearest integer and converts the result to an 
integer variable.
```{r read and clean mrtrash data}
mrtrash_df <- read_excel('Trash-Wheel-Collection-Totals-7-2020-2.xlsx',
                            sheet = 'Mr. Trash Wheel', range = cell_cols('A:N'),
                            skip = 1) %>% 
  janitor::clean_names() %>% 
  rename(dump_num = dumpster) %>% 
  filter(!(is.na(dump_num) | dump_num == 'Grand Total')) %>% 
  mutate(sports_balls= as.integer(round(sports_balls, 0)))

mrtrash_df

colnames(mrtrash_df)
```
This produces a tibble of **453** observations and **14** variables. The key
variables include *dump_num*, *year*, *weight_tons*, *sports_balls*,etc.

The reading and cleaning steps of Professor Trash Wheel sheet are the same.
```{r read and clean proftrash data}
proftrash_df <- read_excel('Trash-Wheel-Collection-Totals-7-2020-2.xlsx',
                            sheet = 'Professor Trash Wheel', 
                           range = cell_cols('A:N'), skip = 1) %>% 
  janitor::clean_names() %>% 
  rename(dump_num = dumpster) %>% 
  filter(!(is.na(dump_num) | dump_num == 'Grand Total')) %>% 
  mutate(sports_balls= as.integer(round(sports_balls, 0)))

proftrash_df
```
This produces a tibble of **71 observations** and **14 variables**.The key
variables include *dump_num*, *year*, *weight_tons*, *sports_balls*,etc.

After getting these 2 datasets, we can join them into 1 tidy dataset. But first,
we can add an additional variable to both datasets before combining to keep 
track of which Trash Wheel is which. 
```{r before table join}
proftrash_df <- proftrash_df %>% 
mutate(label=rep('prof_trash',nrow(proftrash_df)))

proftrash_df

mrtrash_df <- mrtrash_df %>% 
mutate(label=rep('mr_trash',nrow(mrtrash_df))) %>% 
   mutate(dump_num = as.double(dump_num))

mrtrash_df
```
The next step is dataset combination. I would use the function of full_join() to
keep all the data of these 2 datasets, and join them by their shared variables.
```{r table join}
(alltrash_df <- full_join(mrtrash_df,proftrash_df, 
                         by = intersect(colnames(mrtrash_df),
                                        colnames(proftrash_df))))
```
In this way, I get a tibble of **524 observations** and **15 variables**. Simply
speaking, it is like put the proftrash_df at the bottom of mrtrash_df since the
2 dataframes have exactly the same variable names. However, they could also be 
easily distinguished by the different corresponding value of variable "label".

The illustrate the information of the datasets,better, we can extract some of 
the info we want. For example, we can caculate the sum of certain numeric
variables:
```{r sum number}
sum(proftrash_df$weight_tons)

(mrtrash2020_df <- mrtrash_df %>% filter(year == 2020))
sum(mrtrash2020_df$sports_balls)
```
The total weight of trash collected by Professor Trash Wheel is **135.5 tons**.
The total number of sports balls collected by Mr. Trash Wheel in 2020 is **856**.

## Problem 3
This problem uses the FiveThirtyEight data. We’ll use the data in pols-month.csv, unemployment.csv, and snp.csv. Our goal is to merge these into a single data 
frame using year and month as keys across datasets.

First step is to read and clean the pols-month datasets. We use separate() to
break up the variable mon into integer variables year, month, and day; replace 
month number with month name; create a president variable taking values gop and 
dem, and remove prez_dem and prez_gop; and remove the day variable.
```{r pols_df}
(pols_df <- read_csv('fivethirtyeight_datasets/pols-month.csv') %>% 
  separate(mon, into = c('year','month','day'),sep = '-') %>% 
  mutate(year = as.integer(year), month = as.integer(month),
         day=as.integer(day)) %>% 
  mutate(month = month.abb[month]) %>% 
  mutate(president = ifelse(prez_gop == 1, 'gop', 'dem')) %>% 
  select(- c(day, prez_gop, prez_dem)))
```
This produces a tibble of **822 observations** and **9 variables**. The year 
ranges from 1950 to 2015. It includes year-month, gov_gop(the number of 
republican governors on the associated date), sen_gop(the number of republican 
senators on the associated date), rep_gop(number of republican representatives 
on the associated date), gov_dem(the number of democratic governors on the 
associated date), sen_dem(the number of democratic senators on the associated 
date), rep_dem(the number of democratic representatives on the associated date),president(the president was democratic or republican on the associated 
date) .

Second, clean the data in snp.csv using a similar process to the above. For 
consistency across datasets, we arrange according to year and month, and 
organize so that year and month are the leading columns.
Note that again for consistency again datasets, we will replace month number 
with month name.
``` {r snp_df}
(snp_df <- read_csv('fivethirtyeight_datasets/snp.csv') %>% 
  separate(date, into = c('month','day','year'),sep = '/') %>%
  mutate(year = as.numeric(year)) %>%  
  mutate(year = ifelse(year <= 15, 2000+year, 1900+year)) %>% 
  mutate(month = as.numeric(month), day = as.numeric(day)) %>% 
  mutate(month = month.abb[month]) %>% 
  arrange(year, month))
```
This produces a tibble of **787 observations** and **4 variables**. It includes
*year*,*month*,*day*,and *close* (closing values of the S&P stock index on the 
associated date).

Third, we will tidy the unemployment data so that it can be merged with the 
previous datasets. This process will involve switching from “wide” to “long” 
format; ensuring that key variables have the same name; and ensuring that key 
variables take the same values.
```{r unem_df}
(unem_df <- read_csv('fivethirtyeight_datasets/unemployment.csv') %>% 
 pivot_longer(Jan:Dec, names_to = 'month', values_to = 'unemployed_rate') %>% 
  janitor::clean_names())
```
This produces a tibble of **816 observations** and **3 variables**. It includes
*year*, *month*,and the corresponding *unemployment percentage* on that date.

Last, we will join the datasets by merging snp into pols, and merging 
unemployment into the result by using the function left_join().
```{r data merge}
(result_df <- left_join(pols_df, snp_df, by = c('year', 'month')))

(final_result_df <- left_join(result_df, unem_df, by = c('year', 'month')))
```
The *result_df* is a tibble of **822 observations** and **11 variables**, and 
the *final_result_df* is a tibble of **822 observations** and **11 variables**.
Key variables of final result include year-month-day, gov_gop, sen_gop, rep_gop, 
gov_dem, sen_dem, rep_dem, president, close, unemployment_rate.


